---
title: "Homework 1 Solutions"
author: "Jacob Townson"
date: "January 18, 2018"
output:
  pdf_document: default
  html_document: default
---


# 1) Answer the following questions about the dataset "HeightData.txt".

Before we begin, we must read in the text file containing "HeightData.txt". 

```{r}
setwd("C:/Users/obewa_000/Documents/My real documents/University of Louisville/public_work/Classes/Homework/Statistics 668")
D=read.table("HeightData.txt",sep="\t",header=TRUE)
```

##(a - 1pt) Find the average height of all children.

```{r}
mean(D$childHeight)
```


##(b - 1pt) Find the percentage of female children.

```{r}
girls = D$gender=='female'
percent.girls = sum(girls)/length(girls)
percent.girls
```

Thus $48.50107$% of the children are female.

##(c - 1pt) What is the height of the tallest child?

```{r}
max(D$childHeight)
```

Thus the tallest child is $79$ inches tall.

##(d - 1pt) How many families have exactly 4 children?

```{r}
require(tidyr)
require(dplyr)
D.4 = select(D,family,children)
D.4 = distinct(D.4)
D.4 = filter(D.4, children == 4)
length(D.4$family)
```

This is the number of families with exactly $4$ children. It was helpful to reduce the data in this way as now we have a dataframe that contains the exact families with $4$ children as well, which in normal cases could be useful for further analysis.

##(e - 2pts) What are the heights of the children in the family with the tallest mother?

```{r}
require(knitr)
tallmom = max(D$mother)
Dmoms = filter(D, mother == tallmom)
kable(Dmoms)
```

This dataframe shows us all information about the family with the tallest mother.

##(f - 2pts) What percentage of children are taller than both of their parents?

```{r}
total = length(D$childHeight)
Dtall = filter(D, D$childHeight > D$father)
Dtall = filter(Dtall, Dtall$childHeight > Dtall$mother)
tallkids = length(Dtall$childHeight)
percent.tall = (tallkids/total)*100
percent.tall
```

This shows us that $23.8758$% of children are taller than both of their parents.

##(g - 2pts) What is the average height of females whose mother was at least 68 inches tall?

```{r}
Dgals = filter(D, D$mother >= 68)
Dgals = filter(Dgals, Dgals$gender == 'female')
average = mean(Dgals$childHeight)
average
```

Thus we find that the avergae height for girls with mothers taller than $68$ inches is $66.46667$ inches.





# 2)
## (a - 5pts) Suppose that $H_1 ^- ,..., H_m ^-$ are generalized inverses of a matrix $H$ and $\alpha_1,..., \alpha_m$ are scalars. Find a condition on the scalars $\alpha_1,...,\alpha_m$ that makes $\alpha_1 H_1 ^- + ...+ \alpha_m H_m ^-$ a generalized inverse of $H$.

Recall $H^-$ is a generalized inverse of $H$ given that $HH^- H = H$. We know that $H_1 ^- ,..., H_m ^-$ are generalized inverses of a matrix $H$. Thus

$$H(\alpha_1 H_1 ^- + ... + \alpha_m H_m ^-)H = H \alpha_1 H_1 ^- H +...+ H \alpha_m H_m ^- H $$
$$= \alpha_1 H H_1 ^- H +...+ \alpha_m HH_m ^- H = \alpha_1 H +...+ \alpha_m H = (\alpha_1 +...+ \alpha_m)H$$

Thus we can tell that $\alpha_1 H_1 ^- + ...+ \alpha_m H_m ^-$ is a generalized inverse of $H$ if $\alpha_1 +...+ \alpha_m = 1$. QED


## (b - 5pts) Let $H$ be defined as in the homework. Find a generalized inverse of $H$ which has no elements equal to $0$.

  
```{r}
require(MASS)
H = matrix(c(2,2,-1,-1,1,1,3,1,-2), 3,3)
ginv(H)
```






# 3) Consider the large matrix labeled as $O$ in the homework. Use computer software to find the two largest eigenvalues of $O$, denoted by $\lambda_1$ and $\lambda_2$. Letting $u_1$ be the eigenvector of $O$ associated with $\lambda_1$ and $u_2$ be the eigenvector of $O$ associated with $\lambda_2$, compute $A = \lambda_1 u_1 u_1 ' + \lambda_2 u_2 u_2 '$.


  
  
```{r}
O = matrix(c(255,54,71,87,88,93,255,54,255,255,255,255,255,97,71,255,255,255,255,255,101,87,255,255,255,255,255,103,88,255,255,255,255,255,108,93,255,255,255,255,255,115,255,97,101,103,108,115,255),7,7)
image(t(O[7:1,]),col=gray((0:255)/255))
```

Using the $\mathrm{eigen()}$ function, we can find the max eigenvalue and corresponding eigenvectors and calculate $A$ from there.

```{r}
E = eigen(O)
E
E$values[order(E$values)[6:7]]
```

So we can see that the two largest eigenvalues are the first two, meaning that their corresponding eigenvectors are columns 1 and 2 of the matrix containing all of the eigenvectors. 

```{r}
lambda_1 = E$values[1]
lambda_2 = E$values[2]
u_1 = E$vectors[,1]
u_2 = E$vectors[,2]
lambda_1
u_1
lambda_2
u_2
```

These are our largest eigenvalues and their corresponding eigenvectors.

```{r}
A = (lambda_1 * (u_1 %*% t(u_1))) + (lambda_2 * (u_2 %*% t(u_2)))
A
```















```{r echo=FALSE}
setwd("C:/Users/obewa_000/Documents/My real documents/University of Louisville/public_work/")
```


