---
title: "HW5"
author: "Jacob S Townson"
date: "March 26, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1)











#2)










#3)

First we must set up the data. It could have also been read in through a file, however this makes it easier to work from 2 different pc's on the same project.

```{r}
set.seed(123)
x0=1:100/100
x=c(x0,10)
y0=sqrt(1+.1*x0+rnorm(100,sd=.0001))
y=c(y0,0)
```

##a)

Below is a scatterplot of the data, with the line representing our model going through it. As you can see, the outlier at $x=10$ makes the model seemingly strange for the rest of the data. We will address this in later parts of the problem.

```{r fig.height=4}
model3 = lm(y~x)
plot(x,y, main = "scatterplot for #3")
abline(model3$coef)
```


##b)

Below is the residual plot.

```{r fig.height=4}
plot(model3, which = 1, labels.id = '')
```

Below is the studentized residual plot.

```{r fig.height=4}
require(MASS)
plot(studres(model3))
```

Below is the externally studentized residual plot.

```{r fig.height=4}

```


##c)

##d)

##e)








