---
title: "Homework 1 Solutions"
author: "Jacob Townson"
date: "January 17, 2018"
output:
  html_document: default
  pdf_document: default
---


# Answer the following questions about the dataset "HeightData.txt".

Before we begin, we must read in the text file containing "HeightData.txt". 

```{r}
setwd("C:/Users/obewa_000/Documents/My real documents/University of Louisville/public_work/Classes/Homework/Statistics 668")
D=read.table("HeightData.txt",sep="\t",header=TRUE)
```
Now we can work on the required problems.

##(a - 1pt) Find the average height of all children.

```{r}
mean(D$childHeight)
```


##(b - 1pt) Find the percentage of female children.

```{r}
girls = D$gender=='female'
percent.girls = sum(girls)/length(girls)
percent.girls
```


##(c - 1pt) What is the height of the tallest child?

```{r}
max(D$childHeight)
```


##(d - 1pt) How many families have exactly 4 children?

```{r}
require(tidyr)
require(dplyr)
D.4 = select(D,family,children)
D.4 = distinct(D.4)
D.4 = filter(D.4, children == 4)
length(D.4$family)
```

This number contains the number of families with exactly $4$ children. It was helpful to reduce the data in this way as now wehave a dataframe that contains the exact families with $4$ children as well, which in normal cases could be useful for further analysis.

##(e - 2pts) What are the heights of the children in the family with the tallest mother?

```{r}
require(knitr)
tallmom = max(D$mother)
Dmoms = filter(D, mother == tallmom)
kable(Dmoms)
```

This dataframe shows us all information about the family with the tallest mother.

##(f - 2pts) What percentage of children are taller than both of their parents?

```{r}

```


##(g - 2pts) What is the average height of females whose mother was at least 68 inches tall?















```{r echo=FALSE}
setwd("C:/Users/obewa_000/Documents/My real documents/University of Louisville/public_work/")
```


