% TEMPLATE FOR STANDARD QUIZ
\documentclass[11pt,epsfig]{article}

\oddsidemargin=0in
\evensidemargin=0in
\textwidth=6.3in
\topmargin=-0.5in
\textheight=9in

\parindent=0in
\pagestyle{empty}

% C heck if we are compiling under latex or pdflatex, and include the
% appropriate graphics package
\ifx\pdftexversion\undefined
  \usepackage[dvips]{graphicx}
\else
  \usepackage[pdftex]{graphicx}
\fi

\input{testpoints}

\usepackage{amsmath,amsfonts,amssymb,amstext,color,latexsym,graphics,tabulary,tabularx,graphicx,tikz}

\raggedbottom % Makes the bottom margin more flexible (helpful for pictures)

\begin{document}


%%%(change to appropriate class and semester)
Math 105-30 Summer Term III 

%%%(change to appropriate quiz type and date)
Project 1  Due 07/26\hspace{1in} {Name:} {\underline {\hspace{3.5in}}}
\vspace{2pc}

%%%(modify rules, time, points as appropriate)
This project is worth 70 points. You should clearly show all of your work and justify your answers where appropriate. It must be submitted no later than the beginning of class on June 23rd. \textbf{No late projects will be accepted, nor will any e-mail submissions be accepted unless otherwise specified by the instructor.} You may use scrap paper as needed, but please turn in your final project, with complete steps for all of your calculations, on the worksheets that have been provided and make sure that your work is neatly organized and that your final answers are written in complete sentences for all parts. Neatness of your presentation and complete sentences do count! \bigskip

%\noindent You are going to explore the realities of purchasing a home given current market conditions (i.e. current interest rates, salaries, and housing costs). You will be randomly assigned an annual income, an investment account from which you will make a down payment, and a number of bedrooms that your home must have. It will be your responsibility to search for a home that you can afford and satisfies your needs that is currently available for sale according to online listings. You will also need to find current interest rates for mortgages from actual financial institutions. Be careful: there are many different types of mortgages available, and most are more complicated than the mortgages we have studied in class. Additionally, some mortgages are refinance loans, which are not for new homebuyers. If you are in doubt, please feel free to ask questions. You should answer the following questions regarding your home mortgage.

\noindent \textbf{My annual income is $\rule{2cm}{0.15mm}$. My down payment is an investment portfolio in which I invested $\rule{1cm}{0.15mm}$ at the end of every quarter for $\rule{1cm}{0.15mm}$ years at an interest rate of $\rule{1.5cm}{0.15mm}$ compounded quarterly. I need a home with $\rule{1cm}{0.15mm}$ bedrooms.}
\vspace{2pc}

% problem
\begin{problem}{5}
In your own words, describe the difference between the FA formula and the PV formula. Be sure to explicitly describe when you use each. 

\vfill
\end{problem}

% problem
\begin{problem}{5}
For the past ten years your uncle has been depositing \$1,000 at the end of each year in the savings account that pays 6.5\% compounded annually. What was the value of the account just after the tenth deposit? How much interest had been earned?

\vfill
\end{problem}

\newpage

% problem
\begin{problem}{5}
Best Buy offers a TV for \$300 down and \$15 per month for the next 24 months. If interest is charged at 9\% commpounded monthly, find the cash value of the TV. How much total interest will the customer pay with this offer?

\vfill
\end{problem}

% problem
\begin{problem}{5}
A family wants to start a savings account for college expenses 18 years from now, and will make their first deposit at the end of this month. What equal monthly deposits in an account paying 5.0\% compounded monthly are needed in order to accumulate \$25,000 just after the last deposit 18 years from now? How much interest will the account have earned?  

\vfill
\end{problem}

% problem
\begin{problem}{5}
Suppose you purchace a new computer for \$1,500. You may finance the purchase by a loan for one year at an annual interest rate of 15\% compounded monthly. What would be your monthly payment if the loan is amortized? How much interest would you pay?

\vfill
\end{problem}

\showpoints
\end{document}